---
title: "EC1 water quality visualizations"
author: "PR"
date: '2023-02-07'
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = F, message = FALSE, warnings = FALSE)
```


```{r Setup}
# 1. Setup ---------------------------------------------------------------------

## Load packages
require(pacman)
p_load(tidyverse, ## keep things tidy
       cowplot,
       corrplot,
       googledrive) ## download files from google drive

## Set up common columns that are routinely used to in select() and join() calls
common_cols = c("kit_id", "transect_location")

## Set ggplot theme
theme_set(theme_bw())
```

## Thoughts

I don't like Rmd, but I"m doing my best to share my thoughts with my plots since
I'll be out, and this seems like the best format for that. My goal here
is to show what the wq data look like, and explore in some formats that I think 
might be important for visualizing the ions/npoc data. Hopefully something in here
is useful! Please note I'm using L0B here, but the only difference is that flags
are removed for L2, so I'll remove flagged items prior to any analysis. Doing this
cause I already have the code for pulling from the GD, and too lazy to find the L2.


```{r Bring in data, include=FALSE}
## This is KEY: it avoids user input, and will automatically find the token. 
## This has, I think, timed out before. If so, just run drive_auth() and select
## the proper email.
options(gargle_oauth_email = "peter.regier@pnnl.gov")

## Set two paths for the two folders with data we want
data_path <- "https://drive.google.com/drive/folders/1r3sP7tvhG2btZACvxZUdtrtSiLVGUUGp"
metadata_path <- "https://drive.google.com/drive/folders/1IQUq_sD-Jama7ajaZl1zW_9zlWfyCohn"

## Bring in the files you want
file <- drive_ls(data_path) %>% 
  filter(grepl("WaterQuality", name)) %>% 
  pull(name)

metadata <- drive_ls(metadata_path) 


## Now download file to your local
drive_download(file, overwrite = T)
lapply(metadata$name, drive_download, overwrite = T)

## Last, import data
wq_raw <- read_csv(file)

kit_metadata = read_csv(metadata$name[[1]])
collection_metadata = read_csv(metadata$name[[3]])
```


```{r Purge flagged data}

# do it the lazy way
wq_cleaned <- wq_raw %>% 
  filter(is.na(sal_flag)) %>% 
  filter(is.na(ph_flag)) %>% 
  filter(is.na(orp_flag)) %>% 
  filter(is.na(alk_flag)) %>% 
  select(-contains("_flag"))

## Bring in metadata
wq_md <- inner_join(kit_metadata %>% select(kit_id, region), 
                    collection_metadata %>% select(kit_id, water_latitude, water_longitude, 
                                                   water_systemtype, wetland_soil_horizon), 
                    by = "kit_id")

## Join data and metadata
wq <- inner_join(wq_cleaned, wq_md, by = "kit_id")

```

## Distributions

First off, visualizing distributions, our samples skew pretty heavily towards 
fresh/brackish, but we've got a good range from ~0 to seawater. pH is circumneutral, which, sure, don't know what that tells us. ORP is consistently high, and I'm not putting a ton of stock in that, surface waters should be oxic, plus time for equilibration with headspace during transit plus sitting on the titrator waiting to be run = oxygenation. Don't have a point of reference for alkalinity, but seems like a decent spread with values significantly below and above average seawater (100-130 according to google).

```{r}

## Simple helper function
make_density_plot <- function(var){
  
  p0 <- ggplot(wq, aes({{var}})) + 
    geom_histogram(alpha = 0.5)
  
  p1 <-  ggplot(wq, aes(1, {{var}})) + 
    geom_boxplot(alpha = 0.5)
  
  plot_grid(p0, p1, rel_widths = c(1, 0.5))
}

plot_grid(make_density_plot(sal_psu), 
          make_density_plot(ph), 
          make_density_plot(orp_mv), 
          make_density_plot(alk_mgl_caco3))

```

## Correlations

Looking at how these factors relate, the strongest correlation is alkalinity and orp (R = -0.42) followed by alkalinity and pH (R = 0.40). I'm a little surprised by the lack of a relationship between salinity and pH, but maybe that's because there are so many low-salinity samples. 

Interestingly, pretty strong latitudinal and longitudinal correlations. These are likely cause we have two groups: saltier sites at lower lat and higher long, and vice versa. Doesn't quite make sense with pH, but I'd imagine that some of the marshes have super high organics, so lower pH even with higher salinity would make sense. Moral of the story I see here is spatial patterns might be important, but should tease out by region to check.
```{r}
wq %>% 
  select(is.numeric) %>% 
  cor %>% 
  corrplot(method = "number")
```

## Great Lakes correlations

For just the great lakes, really strong positive correlation between salinity and alkalinity, and the lat/long relationships to salinity also held. 

```{r}
wq %>% 
  filter(region == "Great Lakes") %>% 
  select(is.numeric) %>% 
  cor %>% 
  corrplot(method = "number")
```

## Chesapeake Bay correlations

For CB sites, we lose the correlation to longitude, though there's still a pretty clear correlation to latitude (saltier as you move South towards the bay mouth .... duh). Interesting to me is that alkalinity and orp have a strong negative correlation here, and orp doesn't correlate to anything else. Not sure what that means, since there isn't a clear relationship to salinity. maybe alkalinity hints at bedrock composition, which controls infiltration/drainage rates related to redox? That's a stretch...

```{r}
wq %>% 
  filter(region != "Great Lakes") %>% 
  select(is.numeric) %>% 
  cor %>% 
  corrplot(method = "number")
```

## A quick PCA

To understand how water quality parameters relate, and how they differ between regions, I plotted as a PCA. PC1 seems to be alkalinity v ORP, and PC2 seems to be salinity v pH. Interesting that we get clear separation between almost all sites by region along PC2. 

```{r}

## Load relevant packages
p_load(ggfortify,
       ggConvexHull)

## Make PCA
pca1 <- prcomp(wq %>% dplyr::select(sal_psu, ph, orp_mv, alk_mgl_caco3) %>% drop_na(), scale. = TRUE)

## Plot PCA
autoplot(pca1, data = wq, colour = 'region', 
         loadings = TRUE, loadings.label = TRUE, loadings.label.size = 3) +
  geom_convexhull(aes(group = region, fill = region), alpha = 0.2) +
  scale_color_viridis_d() + 
  scale_fill_viridis_d() 

```


## PCA by region

Out of curiosity, how do these relationships differ between region? Not a very appropriate use of PCA, but interesting to see that there are differing relationships by region. I think, depending on what the question is, PCAs by region that show differing relationships might be pretty interesting, and a good way to look at a bunch of different types of data together on one space. 

```{r}
## Make PCA
pca_gl <- prcomp(wq %>% filter(region == "Great Lakes") %>% 
                   dplyr::select(sal_psu, ph, orp_mv, alk_mgl_caco3) %>% drop_na(), scale. = TRUE)

pca_cb <- prcomp(wq %>% filter(region != "Great Lakes") %>% 
                   dplyr::select(sal_psu, ph, orp_mv, alk_mgl_caco3) %>% drop_na(), scale. = TRUE)

## Plot PCA
gl <- autoplot(pca_gl, data = wq %>% filter(region == "Great Lakes"), 
         loadings = TRUE, loadings.label = TRUE, loadings.label.size = 3) + 
  ggtitle("Great Lakes")

cb <- autoplot(pca_cb, data = wq %>% filter(region != "Great Lakes"), 
         loadings = TRUE, loadings.label = TRUE, loadings.label.size = 3) + 
  ggtitle("Chesapeake Bay")

plot_grid(gl, cb, nrow = 1)

```






